---
title: "The Effect of Economic Events on Votes for the President"
author: "Scott Cohn and Samuel Hostetter"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  pdf_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '3'
  toc: yes
header-includes:
    - \usepackage{booktabs}
    - \usepackage{xcolor}
    - \usepackage{hyperref}
    - \hypersetup{}
    - \usepackage{longtable}
    - \usepackage{array}
    - \usepackage{multirow}
    - \usepackage[table]{xcolor}
    - \usepackage{wrapfig}
    - \usepackage{float}
    - \usepackage{colortbl}
    - \usepackage{pdflscape}
    - \usepackage{tabu}
    - \usepackage{threeparttable}
    - \usepackage{threeparttablex}
    - \usepackage[normalem]{ulem}
    - \usepackage{makecell}
    - \usepackage{dcolumn}
biblio-style: apalike
documentclass: article
fontsize: 11
link-citations: yes
bibliography: Biblio.bib
subtitle: 'Resource Economics 312'
abstract: The Presidential Equation is a logistic probability model created by Yale University Professor Ray C. Fair to estimate the Democratic share of votes in any given U.S. presidential election. Fair incorporates both economic and political variables to predict which party will gain the majority of votes. This paper recreates and analyzes the significance of Fair’s popular election model. We then provided a forecast for the 2020 presidential election. We found Fair’s model to hold strong prediction power, providing accurate foresight for every election from 1940 to today. However, the significance of some variables used in his predictions is quite low, leading to questions on how the model remains accurate. 
---
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

```{r Chunk Options, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(stargazer)
library(texreg)
library(ggplot2)
library(ggthemes)
library(tidyverse)
library(summarytools)
library(broom)
library(readxl)
library(kableExtra)
library(car)
library(plotly)
library(lmtest)
opts_chunk$set(echo=FALSE,
               cache=TRUE, autodep=TRUE, cache.comments=FALSE, fig.pos= "h",
               message=FALSE, warning=FALSE)
knitr::read_chunk("/Users/scottcohn/Documents/GitHub/R/PresModels/ModelAnalysis.R")
```
```{r data, message=FALSE, warning=FALSE, cache=TRUE, include=FALSE}
# Import

ModelData <- readxl::read_excel("data.xlsx")
D_Outcome <- readxl::read_excel("data.xlsx", sheet = "Actual Value")

# Trim and Clean
WorkSet <- ModelData[11:35, 1:10]
#WorkSet <- WorkSet[, -WorkSet$rfpred]
names(WorkSet)[names(WorkSet) == 't'] <- 'Year'


dplyr::inner_join(WorkSet, D_Outcome, by = "Year")

WorkSet <-
  data.frame(lapply(WorkSet, function(x)
    as.numeric(as.character(x))))

# Convert VP and actualVP to percentages
WorkSet$VP <- (WorkSet$VP / 100)
WorkSet$actualVP <- (WorkSet$actualVP / 100)

explore_data <- WorkSet[ ,c("Year", "actualVP", "I", "DPER", "DUR", "WAR", "G", "P", "Z")]

I_Recode <- explore_data
I_Recode$DemI <- NA
I_Recode$RepI <- NA
I_Recode$DemI[I_Recode$I == "1"] <- 1
I_Recode$DemI[I_Recode$I != "1"] <- 0
I_Recode$RepI[I_Recode$I == "-1"] <- 1
I_Recode$RepI[I_Recode$I != "-1"] <- 0


model_DemI <- lm(actualVP ~ DemI + DPER + DUR + WAR + G*DemI + P*DemI + Z*DemI, data = I_Recode)
```

# Introduction

The "Presidential Equation" is a logistic probability model created by Yale University Professor Ray C. Fair to explain the Democratic share of votes in any given U.S. presidential election. Fair's model takes into account economic and social deterministic factors that influence voters' decisions. This paper will recreate Fair's equation and use it to develop a forecast for the 2020 presidential election given variable economic conditions. This paper will also briefly compare the Fair model to other presidential forecasting models. 

# Literature Review
The @fair_effect_1978 paper was originally published to create a model broad enough where the prominent voting theories of the time could be tested, and to allow testing of these theories against each other.  The three main theorists Fair cites in his 1978 paper are Anthony Downs (1957), Gerald H. Kramer (1971), and George J. Stigler (1973).

## Anthony Downs
Anthony Downs (1957) establishes a disconnect between voter desires and government desires. In a perfectly informed democracy, every voter should vote for a government that will maximize social welfare, and the government that gets elected should do just that. Realistically, Downs argues, voters and political representatives do not have complete information, resulting in voting patterns that maximize private economic welfare and a government whose goal is to attain "the income, power, and prestige that come with office [@downs_economic]." Fair's proposed relationship between private economic welfare and voting probability was based on the axioms brought forth by Downs.

## Gerald H. Kramer
Gerald Kramer (1971), when he began studying econometrics, was interested the question of how voting behavior was influenced by macroeconomic events. Primarily, he focused on votes for the House of Representatives and Congress. Kramer found that most of the variance in his predictive model was dependent on the change of personal income in the short term. Other variables such as coattails, unemployment, and inflation did not have significant effects. George Stigler (1973) found an error in Kramer's data, and when the experiments were rerun the output showed that inflation has a "modest independent affect" [@rosenthal_gerald_2006]. This effect is looked at in Fair's modified equation. Kramer also argued in his 1971 paper that the effect of presidential elections is much more influenced by non-economic events (i.e. candidate personality) than the elections of the House and Congress. 

## George J. Stigler
George J. Stigler (1973) reviews Kramer's proposed election model and disassembles the significance of Kramer's proposed relationships. Stigler believes voters think about many confounding factors when they vote, not only Kramer's per-capita income belief. He argues that Kramer doesn't account for past experiences of the voter and incorrectly weights recent economic conditions the same as distant conditions [@stigler_general_1973]. Every voter has a different economic past, so grouping all voters together into an average per capita income statistic can lead to erroneous prediction. 

At the time of the original paper, the prevailing theory suggested that a voter looked at the current status and previous performance of the parties seeking power and voted for the party that maximized "future utility" (Kramer 1971, Stigler 1973). A primary assumption, under this theory, was that voters are "self-interested and well-informed" [@fair_effect_1978]. More so, Kramer (1971) suggests the voter votes for the party if their performance is deemed "satisfactory". 

> The standard assumption in [election forecasts] is that voters hold the party that controls the presidency accountable for economic events, rather than, say, the party that controls the Congress (if it is different) or the Board of Governors of the Federal Reserve System [@fair_effect_1978].

Therefore, as theory would suggest, economic events greatly influence the vote for the president in the United States. 

# Data and Methods

## Data
The data are provided by Fair via his website. Included are observations for all of the variables[^1] dating back to 1876. There are observations every four years --- except when presidents have left office early and are subsequently replaced. We have reduced the data to observations later than, and including, 1916. This cut is to account for shifting party idealogies and economic transitions that have taken effect prior to World War I. Given this cut, the data are limited to a meager 25 complete observations. In an already small dataset, this further decreases degrees of freedom and potential modeling power. Upon reciept, the data has been cleaned and tidied. There are no missing values for any of the instances recorded. The data are sourced from the U.S. Bureau of Economic Analysis (BEA) website. Population data are taken from the U.S. Department of Commerce (DOC). Ray Fair has compiled much of the available data to eliminate extraneous variables. The data are readily available on his personal website alongside links to previous updates of the model. 

```{r Fair Data, echo=FALSE, message=FALSE, warning=FALSE}
# Ray Fair Data ---- Insert Tex when done
kable(explore_data,
      booktabs = T,
      format = "latex",
      caption = "Ray Fair Data",
      digits = 2
      ) %>% 
  kable_styling(latex_options = c("striped", "hold_position"))
```

[^1]: See "The Fair Model" section for a detailed description of each variable.

## The Fair Model

The data are presented in Table 1. The updated Fair equation per 1992 is as follows:

\[ V_t = \alpha_1 + \alpha_2 G_t\times I_t + \alpha_3 P_t\times I_t + \alpha_4 Z_t\times I_t + \alpha_5 {DPER_t} + \alpha_6 {DUR_t} +\alpha_7 I_t + \alpha_8 {WAR_t} + \mu_t \]

Where, $\alpha_n$ remain unknown coefficients to be estimated. $I$ denotes incumbancy. $I$ equals 1 if the current president is Democratic, -1 if Republican, 0 otherwise.[^2] $G$ has been modified to represent the real growth rate of GDP per capita for the last three quarters of the election year. In contrast to $G$, a "short horizon" variable, $P$ and $Z$ represent the whole term of the administration and are longer term variables. $P$ represents the absolute value of the growth rate of the GDP deflator for the first fifteen quarters of the current administration. $Z$, the "good news variable", is the number of quarters out of the first fifteen of the current administration where the growth rate of real GDP is greater than 3.2 percent. Fair notes in his construction of the model that psychological research dictates that people will remember extreme events more intensely than normal ones; $Z$ tries to capture the "extreme positive growth outcomes" in accordance with this theory. $DPER$ represents the effect of the current president running while they are in office. $DPER$ equals 1 if current president that will run is Democratic, -1 if Republican, and 0 if the current president will not run while in office. $DUR$ is the duration of the party in office (0 if the current ruling party has been in the White House for one term; $1\times I$ if current party has held office for two terms; $(1+k) \times I$ for three terms; $(1+2k)\times I$ for four terms; and so on, where k is a chosen value of 0.25). $WAR$ is 1 for the election years during and immediately following a U.S. war (1920, 1944 and 1948) and 0 otherwise.

[^2]: $I$ serves two functions in Fair's model. First, it interacts with each respective coeffiecent to flip the sign of economic variables to either increase the chance of Republicans to win (decreasing $V_t$) or increase the chance of Democrats to win (increasing $V_t$). Additionally, $I$ holds a coefficient of its own to demonstrate the effect incumbancy has on voters' decisions.

In forecasts calculated prior to 1996, the original GDP data are presented in a Laspeyres index, which tended to overstate the effect of inflation. Following 1996, GDP is calculated using chain-linked volume series. According to Fair, this more accurately represents the effects of production on the vote-share by providing a better index for growth measurement [@fair_effect_1996]. 

The model above is the most recent iteration of Fair's presedential equation. Table 2 lists the coefficients derived from the data given in this paper. Table 3 below demostrates the prediction power of Fair's model. Appendix A shows the full regression table for all regressions run with different variable combinations. Appendix B shows a graphical analysis of the predicted outcomes using Fair's model versus the actual results.

```{r Fair Coeff, message=FALSE, warning=FALSE, paged.print=FALSE}
simple_model_Fair <- lm(actualVP ~ I + DPER + DUR + WAR + G:I + P:I + Z:I , data = explore_data)
kable(tidy(simple_model_Fair),
      booktabs = T,
      digits = 3,
      format = "latex",
      caption = "Ray Fair 1992 Model Coefficients"
      ) %>% 
  kable_styling(latex_options = c("striped", "hold_position"))
```


```{r Fair VS Actual, echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE}
FairVersusActual <- read_excel("data.xlsx", sheet = "Fair_Versus_Actual_Pres")
kable(FairVersusActual,
      booktabs = T,
      format = "latex",
      caption = "Fair Prediction Compared with Actual Democratic Share of Votes"
      ) %>% 
  kable_styling(latex_options = c("striped", "hold_position"))
```

## Model Replication and Adjustments
### Multicollinearity
Multicollinearity is a common problem within time series data and indicates a high degree of correlation between two or more independent variables. When multicollinearity is present, standard errors inflate and calculated t-values decrease With deflated $t$-values, the probability of a Type II error, or the probability of failing to reject an incorrect hypothesis, increases, impairing the possibility for inference to be made about coefficients. One method of detecting multicollinearity is calculating Variation Inflation Factors ($VIF$) using the formula below.

$$VIF_k = \frac{1}{1 - R^2_k}$$

```{r VIF, message=FALSE, warning=FALSE, echo=FALSE}
tab <- tidy(vif(simple_model_Fair))
kable(tab, 
    booktabs = T,  
    format = "latex",
    digits = 3,
    caption="Variance Inflation Factors for the Ray Fair Regression Model",
    col.names=c("regressor","VIF")) %>% 
  kable_styling(latex_options = c("striped", "hold_position"))
```

The frequently used heuristic for looking at *Variance Inflation Factors* suggests that any *VIF* greater than **10** indicates a problem. The single variable $I$ has a rather high *VIF*, however, this can be attributed to it's interaction affects with the other variables in the model.

### Heteroskedasticity
Heteroskedasticity occurs when residual values are different across different independent variable values. In a homoskedastic model, residuals are of the same magnitude no matter the value of the independent variable. With unequal variances, a model displays incorrect standard errors, which makes inference impossible. To test for heteroskedasticity, we ran a Breusch-Pagan Test (BP Test) at the $0.05$ significance level.

$$\chi^2 = n\times R^2 \sim \chi^2_{(N-1)}$$

```{r}
kable(tidy(bptest(simple_model_Fair)),
    booktabs = T,  
    format = "latex",
    digits = 3,
    caption="Breusch-Pagan Test for Heteroskedasticity"
    )  %>% 
  kable_styling(latex_options = c("striped", "hold_position"))
```

With a calculated $p$-value of $0.574$, we fail to reject the null hypothesis of homoskedasticity for the model. The variance is constant throughout all values of our independent variables.

### Autocorrelation
Autocorrelation occurs in a model when the disturbances influence each other over time. With autocorrelation present, standard errors for each coefficient are wrong and inference would be incorrect. We utilized the Durbin-Watson test to check for autocorrelation within our data. We used the d statistic as calculated below and testing it against the null hypothesis $d=2$ at the $0.05$ significance level.

$$d = \frac{\sum_{t=2}^{T}(e_t-e_{t-1})^2}{\sum_{t=1}^{T}e^2_t}$$

```{r DurbinWatson, message=FALSE, warning=FALSE}
DW <- dwtest(simple_model_Fair)
kable(tidy(DW),
    booktabs = T,  
    format = "latex",
    digits = 3,
    caption="Durbin-Watson Test for Autocorrelation"
    )  %>% 
  kable_styling(latex_options = c("striped", "hold_position"))
```

The Durbin-Watson test returns a Durbin-Watson statistic that is not statistically less than 2. This indicates a lack of positive autocorrelation. We fail to reject the null hypothesis that true autocorrelation is greater than 0.

### Model Significance

A full regression table (Table 10) is available in Appendix A. Specification of the model began with taking all of the available variables and regressing them on the true Democratic share of the vote. 
$$V_t = \alpha_1 + \alpha_2 G_t + \alpha_3 P_t + \alpha_4 Z_t+ \alpha_5 {DPER_t} + \alpha_6 {DUR_t} +\alpha_7 I_t + \alpha_8 {WAR_t} + \mu_t$$

This first model had an $R^2$ of $0.348$ and a calculated $F$-statistic of $1.295$ on 7 and 17 degrees of freedom. The $p$-value for this statistic was $0.3107$. This did not indicate model significance at the 5 percent level. The effects of the parameters were minimal, and the only parameter that appeared significant was the intercept (which was significant at the 1 percent level).

Next, we hypothesized that the party of the incumbent may affect the economic variables and their contribution to the vote-share. The second model (designated as "Interaction" at the top of Table 10 in Appendix A) interacted the variable $I$ with $G$, $P$ and $Z$, while also leaving non-interacted $G$, $P$, and $Z$ in the model.

$$V_t = \alpha_1 + \alpha_2 G_t\times I_t + \alpha_3 P_t\times I_t + \alpha_4 Z_t\times I_t + \alpha_5 {DPER_t} + \alpha_6 {DUR_t} +\alpha_7 I_t + \alpha_8 {WAR_t} + \alpha_9 G_t + \alpha_{10} P_t + \alpha_{11} Z_t + \mu_t$$
Our adjusted-$R^2$ in this model jumped to $0.8155$, with a calculated $F$-statistic of $11.608$ on 10 and 14 degrees of freedom. The $p$-value for this statistic is significant at less than the 1 percent level, indicating that the model is highly significant. In this model, all of the interaction terms are significant at, at least, the 5 percent level. Other significant parameters include the intercept and $DPER$. Given the obvious redundancy of variables, model 3 removes the isolated $G$, $P$ and $Z$.

$$V_t = \alpha_1 + \alpha_2 G_t\times I_t + \alpha_3 P_t\times I_t + \alpha_4 Z_t\times I_t + \alpha_5 {DPER_t} + \alpha_6 {DUR_t} +\alpha_7 I_t + \alpha_8 {WAR_t} + \mu_t$$
The model above is the model used from 1992 to present by Ray Fair. This model has an adjusted-$R^2$ of $0.8327$, with a calculated $F$-statistic of $18.06$. The $p$-value for this statistic is $1.021e-06$, indicating very high significance. Additionally, all of the parameters, except $I$ and $WAR$ have significance at the 5 percent level (unlike $WAR$, $I$ has significant interaction with other parameters). What happens if we remove the $WAR$ variable?

$$V_t = \alpha_1 + \alpha_2 G_t\times I_t + \alpha_3 P_t\times I_t + \alpha_4 Z_t\times I_t + \alpha_5 {DPER_t} + \alpha_6 {DUR_t} +\alpha_7 I_t + \mu_t$$
After removing the $WAR$ variable, the adjusted-$R^2$ goes down to $0.8121$. Many of the parameters become slightly less significant. See Table 10 for a side-by-side comparison of the values. We ran a Joint $F$-test to determine whether the $WAR$ variable is significant.

```{r}
Fair_NoWar <- lm(actualVP ~ I + DPER + DUR + G:I + P:I + Z:I , data = explore_data)
war_anova <- anova(simple_model_Fair, Fair_NoWar)
kable(tidy(war_anova),
    booktabs = T,  
    format = "latex",
    digits = 3,
    caption="Joint F-Test for WAR parameter"
    )  %>% 
  kable_styling(latex_options = c("striped", "hold_position"))
```
The results of the Joint $F$-test indicate failure to reject the null hypothesis that the two models are equivalent. Given the higher adj-$R^2$ and higher signficance of the individual parameters, our model will continue to utilize the $WAR$ variable. 

Thus, the final model is: 
$$V_t = \alpha_1 + \alpha_2 G_t\times I_t + \alpha_3 P_t\times I_t + \alpha_4 Z_t\times I_t + \alpha_5 {DPER_t} + \alpha_6 {DUR_t} +\alpha_7 I_t + \alpha_8 {WAR_t} + \mu_t$$

# Forecasting
A key adjunct to this paper is a presidential vote forecast. Much of the notoriety surrounding Fair’s presidential equation is it’s accuracy in predicting the outcome of elections. In 2014, Fair constructed a forecast to the 2016 election by setting all non-economic values to fixed values and placing predictions on economic variables (Fair 2014). The three separate forecasts for a booming, continuous, or sluggish economy provided by Fair indicated a Republican win barring an economic boom. The economic conditions that followed the 2014 paper were closer to his sluggish scenario, affirming his preliminary forecast. Fair's forecast is displayed in Table 7 below.

To further Fair’s work, this paper will conduct a forecast for the 2020 presidential election. The $G$, $P$, $Z$, $DPER$, $DUR$, and $I$ variables hold different values than the 2014 Fair forecast and are as follow:

The non-economic variables in all three scenarios are fixed. $I$ = $-1$ (Republican party is in power), $DPER$ = $-1$ (assuming incumbent president will run again), $DUR$ = 0 (Republican party has been in power for only one term), and $WAR$ = 0. In the “continued economic conditions” scenario, we inputted the growth rate of per capita GDP ($G$) as the growth rate experienced at the end of 2017 (2017:3 - 2017:4) at the annual rate. We did not add anything to the assumptions Fair makes when predicting GDP. We make the assumption that his predictions are sound, and thus we are excluding discussion on those formulations from this paper. The value of 1.85 comes from data taken from the Federal Reserve Bank of St. Louis [@FRED_1947]. The absolute value of the GDP deflator ($P$) was calculated similarly to the per capita GDP growth rate, taking economic data from 2017 (2016:4 - 2017:4) and determining the growth rate of inflation at the annual rate. For the same reason we are excluding discussion of the derivation of $G$, we do the same for $P$. It is explained in depth in Fair 2014. The calculated value for $P$ was 2.35 [@us_bureau_of_economic_analysis_gross_1947]. Typically, foreasting with this model occurs using the first 8 quarters of the current administration, but there have not been 8 quarters yet. Given there haven't been enough “good news” quarters ($Z$) yet, the data are showing an upward trend in GDP per capita growth rate. With this trend, our estimate of $Z$ for the “continued economic condition” forecast is 4.

The remaining two scenarios have variable values scaled up or down according to our estimates. The three scenarios give a good forecast of the presidential vote count in 2020 based on three variable, economic conditions. The next step is to run the three separate scenarios through Fair's presidential equation, with the coefficients shown in the first column of Table 2. In Table 8, the results of the forecast are shown. In all three scenarios, the Democratic share of votes is under 50 percent, indicating a very high liklihood of a Republican victory in the 2020 election. 

```{r}
Fair2014Forecast <- read_excel("data.xlsx", sheet = "Fair2014Forecast")
kable(Fair2014Forecast,
    booktabs = T,  
    format = "latex",
    digits = 2,
    caption="2016 Election Forecast"
    )  %>% 
  kable_styling(latex_options = c("striped", "hold_position"))

Forecast2020 <- read_excel("data.xlsx", sheet = "Forecast")
kable(Forecast2020,
    booktabs = T,  
    format = "latex",
    digits = 2,
    caption="2020 Election Forecast"
    )  %>% 
  kable_styling(latex_options = c("striped", "hold_position"))
```


# Other Presidential Forecasting Models
Fair's model is notorious for it's size and number of variables. As election forecasting models have grown in popularity, some of the emerging models take similiar approaches to Fair. Others, notably @campbell_forecasting_1992, opt for a much tighter model specification with fewer independent variables to gain similiar accuracy in their estimates.

Other popular general election models (like the polls-plus, polls-only and now-cast from FiveThirtyEight) combine machine learning algorithms with regression analysis to make estimations. These models combine poll results[^3] with economic data, and run upwards of tens of thouseands of simulations once the models are calibrated to their specifications [@silver_users_2016]. Many of these models are strong in their predictions. They are less robust at determining the precise effect of each variable on the outcome. Often a combination of machine learning and classical regression techniques in these models yield the best results.

[^3]: "The idea behind an election forecast like FiveThirtyEight’s is to take polls ('Clinton is ahead by 3 points') and transform them into probabilities ('She has a 70 percent chance of winning')" [@silver_users_2016].

# Discussion

The incumbent variable ($I$) in Fair's model imposes an equality constraint. When interpreting the interaction coefficients in front of $G$, $P$, and $Z$, an assumption is made that the effect of a one unit change in these variables is identical for when $I$ equals $1$ or $-1$. Or, a change in economic conditions has the same effect on voter preference no matter which party is currently in office. The assumption of equal effects can be naive, for the two parties are known to have unequal effects on the economy.

To test the validity of the restriction, we ran Fair’s model with the incumbency variable split into two groups: Democrats and Republicans. To satisfy these conditions, we removed $I$ and replaced it with binary variable $DemI$, which equals $1$ when a democratic president is in office and $0$ when a Republican president is in office. The split allows for two sets of coefficients to be analyzed: one when a Democrat is in office and one when a Republican is in office. Below are the coefficients and Variation Inflation Factors. 

$$V_t = \alpha_1 + \alpha_2 G_t\times DemI_t + \alpha_3 P_t\times DemI_t + \alpha_4 Z_t\times DemI_t + \alpha_5 {DPER_t} + \alpha_6 {DUR_t} +\alpha_7 DemI_t + \alpha_8 {WAR_t} + \alpha_9 G_t + \alpha_{10} P_t + \alpha_{11} Z_t + \mu_t$$

```{r Dem_Coeff, echo=FALSE, message=FALSE, warning=FALSE}
kable(tidy(model_DemI),
      booktabs = T,
      digits = 3,
      format = "latex",
      caption = "Split Incumbent Coefficents"
      ) %>% 
  kable_styling(latex_options = c("striped", "hold_position"))
```

```{r Dem_Vif, echo=FALSE, message=FALSE, warning=FALSE}
Dem_vif <- tidy(vif(model_DemI))
kable(Dem_vif, 
    booktabs = T,  
    format = "latex",
    digits = 3,
    caption="Variance Inflation Factors for Split Incumbent",
    col.names=c("regressor","VIF")) %>% 
  kable_styling(latex_options = c("striped", "hold_position"))
```
The “Split Incumbent” model brings party-specific insights on the effect economic factors have on voter preferences. The adj-$R^2$ for Fair’s equal effect model is $0.833$ (Table 12) compared to our split incumbent model’s adj-$R^2$ value of $0.815$ (Table 13). The similar strength of the models show that splitting the incumbent interaction term does not adversely affect the significance of the model. 

Our model brings back the $G$, $P$, and $Z$ independent variables that Fair removed. In the Split Incumbent model, the coefficients in front of these non-interacted terms indicate the effect of a one unit change in the economic condition has on the democratic share of votes when a Republican is in office. The coefficients are then changed by the interaction term when a Democratic president is in office.

If Fair’s model is correct, we are to expect equal but opposite coefficients for when Republicans or Democrats are in office. However, the economic variable $P$ has significantly different effects on voter behavior depending on who is in office. When a Republican is in office, a one percent increase in the absolute value of the growth rate of the GDP deflator increases the Democratic share of the two-party vote by $0.006$ percent. When a Democrat is in office, a one percent increase in the absolute value of the growth rate of the GDP deflator decreases the Democratic share by $0.013$. With such a significant difference of effects, the Split Incumbent model illuminates a possible flaw within Fair’s model.

# Conclusion
Many traditional journalists conduct revisionist history when talking about how an outcome happened. On the flip side, data journalism and election forecasting disrupt the traditional outlets of political media. These models circumvent the political hype surrounding election predictions and instead focus on concrete economic and politcal data. Often, correcteness is praised higher than model confidence. Regardless of how well specified the model is, an incorrect forecast is still an incorrect forecast. Traditional media does not care about specificiation bias --- they care about who will win the Presidency, and that's it.

Problems often arise in interpretations of these models. These forecasts and model predictions are representations of uncertainty. To the untrained eye, these forecasts can seem more absolute than they are. Using the example of a presidential election, if the model has a point estimate of 52 percent of the votes going towards the Republican nominee there can be a confidence interval that spans a 49 percent outcome to a 55 percent outcome. Both ends of this interval *can be* equally likely. This paper brings a scrutinizing eye to one of the most accurate election equations: the Fair model. Through trying find common errors in time series analysis (including autocorrelation and heteroskedasticity), we explored different possible weaknesses in Fair's model in the explaining the accuracy of the popular model. Many of our attempts came up fruitless. Even when run through a gauntlet of specification tests, the Fair model remains robust.

The forecast section of this paper provides insight on how the Fair model can be used to give preliminary estimates for upcoming elections. A key distinction between the forecast and the model is a decrease in confidence. The equation Fair uses to generate his steadfast predictions relies on data only available during the current election year. The forecast in this paper generates three separate estimates for three of the variables ($G$,$P$, and $Z$) based on different possible economic scenarios leading up to the 2020 election. It is very possible that none of the exact scenarios pan out in the coming years. In time, the forecast's strength can be tried with actual economic variables and actual election results. Fair's latest forecast in 2014 correctly forecasted a Republican win in 2016 with respect to actual economic conditions.

\clearpage
# Appendix A --- Regression Results
```{r RegTables, message=FALSE, warning=FALSE}
# TABLE SOMEWHERE WITH REGRESSIONS
# simple_model <- lm(actualVP ~ I + DPER + DUR + WAR + G + P + Z, data = explore_data )
# simple_model_2 <- lm(actualVP ~ I + DPER + DUR + WAR + G*I + P*I + Z*I , data = explore_data )
# simple_model_4 <- lm(actualVP ~ I + DPER + DUR + G:I + P:I + Z:I + WAR, data = explore_data )
# simple_model_3 <- lm(actualVP ~ I + DPER + DUR + G:I + P:I + Z:I , data = explore_data )
# 
# stargazer(simple_model, simple_model_2, simple_model_Fair, simple_model_3,
#           title = "Regression Results",
#           column.labels=c("Full","Interaction", "Fair", "No WAR"),
#           type = "latex",
#           digits = 3,
#           align=TRUE,
#           style = "aer",
#           header = FALSE
#           )
```

\begin{table}[!htbp] \centering 
  \caption{Regression Results} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lD{.}{.}{-3} D{.}{.}{-3} D{.}{.}{-3} D{.}{.}{-3} } 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
\\[-1.8ex] & \multicolumn{4}{c}{actualVP} \\ 
 & \multicolumn{1}{c}{Full} & \multicolumn{1}{c}{Interaction} & \multicolumn{1}{c}{Fair} & \multicolumn{1}{c}{No WAR} \\ 
\\[-1.8ex] & \multicolumn{1}{c}{(1)} & \multicolumn{1}{c}{(2)} & \multicolumn{1}{c}{(3)} & \multicolumn{1}{c}{(4)}\\ 
\hline \\[-1.8ex] 
 I & 0.017 & -0.020 & -0.023 & -0.008 \\ 
  & (0.043) & (0.028) & (0.024) & (0.024) \\ 
  & & & & \\ 
 DPER & 0.037 & 0.047^{**} & 0.045^{***} & 0.051^{***} \\ 
  & (0.042) & (0.019) & (0.015) & (0.016) \\ 
  & & & & \\ 
 DUR & -0.042 & -0.028 & -0.027^{*} & -0.020 \\ 
  & (0.036) & (0.016) & (0.013) & (0.013) \\ 
  & & & & \\ 
 WAR & 0.018 & 0.035 & 0.050^{*} &  \\ 
  & (0.073) & (0.039) & (0.028) &  \\ 
  & & & & \\ 
 G & -0.001 & -0.001 &  &  \\ 
  & (0.003) & (0.002) &  &  \\ 
  & & & & \\ 
 P & -0.005 & -0.004 &  &  \\ 
  & (0.007) & (0.003) &  &  \\ 
  & & & & \\ 
 Z & 0.006 & 0.000 &  &  \\ 
  & (0.006) & (0.003) &  &  \\ 
  & & & & \\ 
 I:G &  & 0.007^{***} & 0.007^{***} & 0.007^{***} \\ 
  &  & (0.001) & (0.001) & (0.001) \\ 
  & & & & \\ 
 I:P &  & -0.010^{**} & -0.009^{**} & -0.011^{***} \\ 
  &  & (0.003) & (0.003) & (0.003) \\ 
  & & & & \\ 
 I:Z &  & 0.008^{**} & 0.008^{***} & 0.006^{**} \\ 
  &  & (0.003) & (0.003) & (0.002) \\ 
  & & & & \\ 
 Constant & 0.467^{***} & 0.473^{***} & 0.462^{***} & 0.466^{***} \\ 
  & (0.045) & (0.021) & (0.007) & (0.007) \\ 
  & & & & \\ 
Observations & \multicolumn{1}{c}{25} & \multicolumn{1}{c}{25} & \multicolumn{1}{c}{25} & \multicolumn{1}{c}{25} \\ 
R$^{2}$ & \multicolumn{1}{c}{0.348} & \multicolumn{1}{c}{0.892} & \multicolumn{1}{c}{0.881} & \multicolumn{1}{c}{0.859} \\ 
Adjusted R$^{2}$ & \multicolumn{1}{c}{0.079} & \multicolumn{1}{c}{0.815} & \multicolumn{1}{c}{0.833} & \multicolumn{1}{c}{0.812} \\ 
Residual Std. Error & \multicolumn{1}{c}{0.067 (df = 17)} & \multicolumn{1}{c}{0.030 (df = 14)} & \multicolumn{1}{c}{0.029 (df = 17)} & \multicolumn{1}{c}{0.030 (df = 18)} \\ 
F Statistic & \multicolumn{1}{c}{1.295 (df = 7; 17)} & \multicolumn{1}{c}{11.608$^{***}$ (df = 10; 14)} & \multicolumn{1}{c}{18.060$^{***}$ (df = 7; 17)} & \multicolumn{1}{c}{18.286$^{***}$ (df = 6; 18)} \\ 
\hline \\[-1.8ex] 
\textit{Notes:} & \multicolumn{4}{l}{$^{***}$Significant at the 1 percent level.} \\ 
 & \multicolumn{4}{l}{$^{**}$Significant at the 5 percent level.} \\ 
 & \multicolumn{4}{l}{$^{*}$Significant at the 10 percent level.} \\ 
\end{tabular} 
\end{table} 

```{r RegTables_Split, message=FALSE, warning=FALSE}
# model_DemI <- lm(actualVP ~ DemI + DPER + DUR + WAR + G*DemI + P*DemI + Z*DemI, data = I_Recode)
# 
# stargazer(model_DemI,
#           title = "Regression Results",
#           column.labels=c("Split Incumbent"),
#           type = "latex",
#           digits = 3,
#           align=TRUE,
#           style = "aer",
#           header = FALSE
#           )
```
\begin{table}[!htbp] \centering 
  \caption{Regression Results} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lD{.}{.}{-3} } 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
\\[-1.8ex] & \multicolumn{1}{c}{actualVP} \\ 
 & \multicolumn{1}{c}{Split Incumbent} \\ 
\hline \\[-1.8ex] 
 DemI & -0.041 \\ 
  & (0.057) \\ 
  & \\ 
 DPER & 0.047^{**} \\ 
  & (0.019) \\ 
  & \\ 
 DUR & -0.028 \\ 
  & (0.016) \\ 
  & \\ 
 WAR & 0.035 \\ 
  & (0.039) \\ 
  & \\ 
 G & -0.008^{***} \\ 
  & (0.002) \\ 
  & \\ 
 P & 0.006 \\ 
  & (0.005) \\ 
  & \\ 
 Z & -0.007^{*} \\ 
  & (0.004) \\ 
  & \\ 
 DemI:G & 0.014^{***} \\ 
  & (0.003) \\ 
  & \\ 
 DemI:P & -0.019^{**} \\ 
  & (0.007) \\ 
  & \\ 
 DemI:Z & 0.016^{**} \\ 
  & (0.006) \\ 
  & \\ 
 Constant & 0.494^{***} \\ 
  & (0.033) \\ 
  & \\ 
Observations & \multicolumn{1}{c}{25} \\ 
R$^{2}$ & \multicolumn{1}{c}{0.892} \\ 
Adjusted R$^{2}$ & \multicolumn{1}{c}{0.815} \\ 
Residual Std. Error & \multicolumn{1}{c}{0.030 (df = 14)} \\ 
F Statistic & \multicolumn{1}{c}{11.608$^{***}$ (df = 10; 14)} \\ 
\hline \\[-1.8ex] 
\textit{Notes:} & \multicolumn{1}{l}{$^{***}$Significant at the 1 percent level.} \\ 
 & \multicolumn{1}{l}{$^{**}$Significant at the 5 percent level.} \\ 
 & \multicolumn{1}{l}{$^{*}$Significant at the 10 percent level.} \\ 
\end{tabular} 
\end{table} 

# Appendix B --- Graph
```{r, out.width = '65%', fig.align = "center"}
library(ggthemes)
Fair_Actual_Graph <- ggplot(FairVersusActual, aes(x = FairVersusActual$Year)) +
  geom_line(aes(y = FairVersusActual$Fair_Prediction, color = "Fair Prediction")) +
  geom_line(aes(y = FairVersusActual$Actual_Vote_Share, color = "Actual Vote Percentage")) +
  geom_line(aes(y = FairVersusActual$Error, color = "Error")) +
  ggtitle("Democratic Share of Presidential Vote: Fair versus Actual") +
  xlab("Year") +
  ylab("Vote Percentage")


Fair_Actual_Graph

```

# Technical Documentation
Written using R [@R].

Packages used:

* knitr, @R-knitr
* MASS, @MASS
* xtable, @xtable
* mosaic, @mosaic
* readxl, @readxl
* dplyr, @dplyr
* Stargazer, @stargazer
* DataExplorer, @DataExplorer
* tidyverse, @tidyverse
* randomForest, @randomForest
* car, @car
* gplot2, @ggplot2
* sjPlot, @sjPlot
* broom, @broom
* texreg, @texreg
* forecast, @forecast
* lmtest, @lmtest


# References
